Stability of variational quantum dynamics with Gaussian approximants

### 2013-11-07 paper.tex
writing

\def\G{|Γ〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}

@@

The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  To test how well a wave function approximates the dynamics of a system, the difference between the two sides of Schrödinger's equation can be measured by the Hilbert space norm.  It remains to choose a class of approximants.  The gaussian wave packets are an common choice, because they differ in their scale and their central position and momentum, useful properties to determine for any quantum state.

@@

@@

Expanding a field over Gaussian functions is a rather obvious idea, and physicists are not the only people to have thought of it.  The mathematicians and signal theorists who developed wavelet expansions have also analysed Gaussians.  They call this type of expansion a Gabor frame, or a windowed Fourier transform.  They have prooved several results that are not well known in physics: for example, that discrete sets of coherent states are complete, not just the continuum.  They have also studied overcomplete sets of vectors in general, under the field of frame theory.  Many of these general ideas insight into the specific case of coherent state frames.

@@

Consider a linear combination of kets,
$$|ψ〉=c₁|a₁〉+c₂|a₂〉+⋯+c_R|a_R〉.$$
This can be written as a matrix product $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}^{\rm T}$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.  Alternatively, the kets $|a_r〉$ can be regarded as infinite dimensional columns, and $|A〉$ as a matrix, representing a linear operator from ${\bf C}^R$ into ket space.  This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}^{\rm T}$, that maps kets back to column vectors.  Under the rules of matrix multiplication, the product
$$〈A|A〉=\pmatrix{〈a₁|a₁〉&&〈a₁|a_R〉\cr &\ddots&\cr 〈a_R|a₁〉&&〈a_R|a_R〉}$$
is an operator on ${\bf C}^R$, and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ is a Hilbert space operator.  Other matrix products work too; provided the $|a_r〉$ are linearly independent, $|A〉$ will have a singular value decomposition $|A〉=|U〉SV†$, where $|U〉$ is a row of $R$ kets, $S$ is an $R×R$ diagonal matrix of complex numbers, and $V$ is an orthogonal $R×R$ complex matrix%
\footnote{${}^1$}{In fact, $|A〉$ has an SVD regardless, but $|U〉$ and $S$ are smaller.  When the $|a_r〉$ are linearly dependent, they span a space of dimension less than $R$, and there is no way to choose in which directions it should be extended to give $R$ orthonormal kets.  In this case, $|U〉$ will be a row of ${\rm rank}|A〉$ kets, and $S$ will be ${\rm rank}|A〉×R$.}.
In this paper, things written as kets have a finite number of infinite dimensional columns, and bras are the other way round.

Even more general matrices are sometimes useful.  For instance, the operator
$$M=\pmatrix{0&|A〉\cr 〈A|&0}$$
can be applied to columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}^{\rm T}$, giving objects of the same form.  The top left zero is an operator on Hilbert space, while the bottom right one is an $R×R$ matrix.  The general rule is that the things in each row must have the same column dimension, and visa-versa.  Things that have finite dimensional rows and columns will be written as matrices, and things that are infinite in both directions as Hilbert space operators.  These are both just written as letters.

@@

Schrödinger's equation is very simple.  The main challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand this over a regular basis grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes mere thousands of particles.  In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
$$|Ψ(t)〉=|ψ(z(t))〉,$$
or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\pmatrix{z₁(t)&…&z_R(t)}^{\rm T}$ is a vector of complex parameters.  This poses two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.

@@

If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
$$i\hbar|Dψ(z)〉Dz=H|ψ(z)〉.$$
On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}^{\rm T}$.  This equation is unlikely to have an exact a solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to lie outside this subspace.  In other words, the restricted form of the approximation does not include the solution to Schrödinger's equation.  We can, however, look for a least-squares solution, a vector $\dot z(z)$ that minimises the residual
$$\bigl\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\bigr\|²,\tag\residual$$
and solve the ordinary differential equation
$$Dz(t)=\dot z(z(t)),\tag\ode$$
giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.  When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.

@@

The approximant studied in this paper has the form
$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
	=|A〉d,\tag\Ad$$
where $A=\pmatrix{α₁&⋯&α_R}^{\rm T}$ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.  The form on the left is more complicated than the $|A〉d$ one, but it is an analytic function of the parameters, which is convenient for derivatives and optimisation.  Wavelet theorists have studyed the conditions on the set of ampltiudes $A$ that allow any state to be expanded in this form, and the stability of expansions as the density of ampltiudes decreases towards the minimum for completeness.  This will be discussed later.

@@

The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.  We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.  We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.  Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}^{\rm T},$
such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.

@@

There is another way to derive coefficients of the constant function, which might be more stable.  Suppose that $D$ is a subset of the real line, large enough that the tails of gaussians are neglibible outside it.  Let 
$$U=π^{-½}\pmatrix{e^{-(x-x₁)²}&⋯&e^{-(x-x_R)²}},$$
and $w$ be a column vector of $R$ real weights.  Then define a residual
$$r(w)=\int_D(Uw-1)².$$
The derivative of this with respect to the weights is
$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U^{\rm T}Uw-U.$$
(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c^{\rm T}ab$.)
Unfortunately,
$$\int (U^{\rm T}U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
	={1\over\sqrt{2π}}e^{(x_i-x_j)²/2},$$
and $\int u_i=1$, so this is the identical problem again.

@@

To prevent the weights from oscillating in order to fit the edges of the samples, it is tempting to do a weighted least squares fit, which emphasises fitting the centre of the cluster over fitting the edges.  There are two ways to do this.  Firstly, instead of expanding the constant function $f(x)=1$, we can expand a peaked function such as $f(x)=e^{-x²}$, then divide the coefficient of $e^{-(x-x_r)²}$ by $e^{-(x_r)²}$ to get a weight.  This mimics the expansions we need to do in practice, where the sample points cover the region where the expanded function has large values.  The residual in this case is
$${\cal E} = \int\left(e^{-x²}-∑_rd_re^{(x-x_r)²}\right)²;$$
the only change to the normal equations is that the right hand side $\pmatrix{1&⋯&1}^{\rm T}$ is replaced by a constant multiple of 
$$b=\pmatrix{e^{-x₁²/2}&⋯&e^{-x_R²/2}}.\tag\rhs$$
For now, I'm ignoring factors of 2, $\sqrt π $ and so on: how the weights oscillate is more interesting than their size.  This seems to work: with random sample points, the weights still oscillate, but remain within an order of magnitude or two, where the weights from expanding the constant function vary by a factor of a million.

>>>>

\def\G{|Γ〉}
\def\Gx{|Γ_χ〉}
\def\cite#1{$[\hbox{\tt #1}]$}
\def\T{^{\rm T}}

@@

The simplest mechanical systems allow Schrödinger's law of motion to be solved in closed form, but most quantum problems must be solved approximately.  This can be done variationally, specifying a class of approximants that can be manipulated exactly, then finding one of them that most nearly solves the problem.  To measure how well a wave function $|ψ(t)〉$ approximates the dynamics of a system, we will compare the left and right sides of Schrödinger's equation, $i\hbar{d|ψ〉\over dt}$ and $H|ψ〉$; these are both vectors in Hilbert space, and the norm defines a distance between them.  It remains to choose a class of approximants.  The gaussian wave packets are an common choice, because they differ in their scale and their central position and momentum, useful properties to determine for any quantum state.

@@

Both condition and numerical stability refer to reasons that a computer can give the wrong answer to a numerical problem.  The distinction between them is a bit subtle.  A problem is ill-conditioned if the solution is very sensitive to the data with which the problem is stated: the canonical example is computing $\sin(x)$ for $x=10⁶$, whence a relative error of $π×10^{-6}$ in $x$ causes a 100\% error in $\sin(x)$.  The key thing to note is that this has nothing to do with how $\sin(x)$ is computed—the error in $x$ could be due to measurement uncertainty, in which case there would be no way to calculate $\sin(x)$: we'd need to measure it directly.  Condition is a property of problems.  Stability is a property of solutions.  For example, if we asked a computer to calculate $1-\cos(x)$ in the obvious way, with $x$ between 0 and $10^{-10}$, the result might vary randomly around 0, in a range determined by the precision with which the computer stored numbers.  But this problem is not intrinsically hard: at any reasonable level of precision, it amounts to finding $x²/2$.  The method of subtracting a numerical value of $\cos(x)$ from 1, which generates random numbers, is unstable.

@@

Expanding a field over Gaussian functions is a rather obvious idea, and physicists are not the only people to have thought of it.  The mathematicians and signal theorists who developed wavelet expansions have also analysed Gaussians.  They call this type of expansion a Gabor frame, or a windowed Fourier transform.  They have proved several results that are not well known in physics: for example, that discrete sets of coherent states are complete, not just the continuum.  They have also studied overcomplete sets of vectors in general, under the field of frame theory.  Many of these general ideas insight into the specific case of coherent state frames.

@@

Consider a linear combination of kets,
$$|ψ〉=c₁|a₁〉+c₂|a₂〉+⋯+c_R|a_R〉.$$
This can be written as a matrix product $|ψ〉=|A〉c$, where $c=\pmatrix{c₁&⋯&c_R}\T$ is a column vector of coefficients, and $|A〉=\pmatrix{|a₁〉&⋯&|a_R〉}$ is a row vector of kets.  Alternatively, the kets $|a_r〉$ can be regarded as infinite dimensional columns, and $|A〉$ as a matrix, representing a linear operator from ${\bf C}^R$ into ket space.  This operator has an adjoint, $〈A|=\pmatrix{〈a₁|&⋯&〈a_R|}\T $, that maps kets back to column vectors.  Under the rules of matrix multiplication, the product
$$〈A|A〉=\pmatrix{〈a₁|a₁〉&&〈a₁|a_R〉\cr &\ddots&\cr 〈a_R|a₁〉&&〈a_R|a_R〉}$$
operates on ${\bf C}^R$, and $|A〉〈A|=|a₁〉〈a₁|+⋯+|a_R〉〈a_R|$ operates on Hilbert space.  Other matrix products work too; provided the $|a_r〉$ are linearly independent, $|A〉$ will have a singular value decomposition $|A〉=|U〉SV†$, where $|U〉$ is a row of $R$ kets, $S$ is an $R×R$ diagonal matrix of complex numbers, and $V$ is an orthogonal $R×R$ complex matrix%
\footnote{${}^1$}{In fact, $|A〉$ has an SVD regardless, but $|U〉$ and $S$ are smaller.  When the $|a_r〉$ are linearly dependent, they span a space of dimension less than $R$, and there is no way to choose in which directions it should be extended to give $R$ orthonormal kets.  In this case, $|U〉$ will be a row of ${\rm rank}|A〉$ kets, and $S$ will be ${\rm rank}|A〉×R$.}.
In this paper, things written as kets have a finite number of infinite dimensional columns, and bras are the other way round.

Even more general matrices are sometimes useful.  For instance, the operator
$$M=\pmatrix{0&|A〉\cr 〈A|&0}$$
acts on columns of the form $\pmatrix{|ψ〉&z₁&⋯&z_R}\T $ to give columns of the same form.  The top left zero is an operator on Hilbert space, while the bottom right one is an $R×R$ matrix.  The general rule is that the things in each row must have the same column dimension, and visa-versa.  Things that have finite dimensional rows and columns will be written as matrices, and things that are infinite in both directions as Hilbert space operators.  These are both just written as letters.

@@

Schrödinger's equation is very simple.  The main challenge of solving it numerically lies in representing the ket $|Ψ(t)〉$—as is well known, the storage required to expand this over a regular basis grows exponentially with the number of particles being simulated, and exceeds the number of atoms in the known universe when the system includes mere thousands of particles.  In order to compute dynamics by propagating a quantum state, we must restrict the form of $|Ψ〉$, and require that
$$|Ψ(t)〉=|ψ(z(t))〉,$$
or $|Ψ〉=|ψ\circ z〉$, where $z(t)=\pmatrix{z₁(t)&…&z_R(t)}\T $ is a vector of complex parameters.  This poses two challenges: to find a form $|ψ〉$ that can approximate the states $|Ψ(t)〉$ with a reasonable number of parameters $R$, and to stably and efficiently compute $z(t)$ to accomplish that approximation.

@@

If the approximant $|Ψ(t)〉=|ψ(z(t))〉$ is substituted in Schrödinger's equation, and the chain rule applied, the result is
$$i\hbar|Dψ(z)〉Dz=H|ψ(z)〉.$$
On the left hand side, the kets in the row vector $|Dψ(z(t))〉$ are combined with coefficients from the column vector $Dz(t)=\pmatrix{Dz₁(t)&…&Dz_R(t)}\T $.  This equation is unlikely to have an exact solution: the kets $|∂_rψ(z(t))〉$ span an $R$-dimensional subspace of Hilbert space, and the ket $H|ψ(z(t))〉$ is likely to lie outside this subspace.  In other words, the restricted form of the approximation does not include the exact solution to Schrödinger's equation.  We can, however, look for a least-squares solution, a vector $\dot z(z)$ that minimises the residual
$$\bigl\|i\hbar|Dψ(z)〉\dot z(z)-H|ψ(z)〉\bigr\|²,\tag\residual$$
and solve the ordinary differential equation
$$Dz(t)=\dot z(z(t)),\tag\ode$$
giving a state $|ψ(z(t))〉$ that hopefully approximates the solution to Schrödinger's equation.  When the norm used is the one derived from the inner product of Hilbert space, this is called the Dirac-Frenkel method.

@@

The approximant studied in this paper has the form
$$|ψ(φ₁,α₁,⋯,φ_R,α_R)〉=∑_{r=1}^R e^{φ_r+α_ra†}|0〉
	=\pmatrix{|α₁〉&⋯&|α_R〉}\pmatrix{e^{φ₁+½|α₁|²}\cr\vdots\cr e^{φ_R+½|α_R|²}}
	=|A〉d,\tag\Ad$$
where $A=\pmatrix{α₁&⋯&α_R}\T $ is a vector of complex amplitudes, $|α_r〉$ is a coherent state, and $φ_r$ is a complex weight.  The form on the left is more complicated than the $|A〉d$ form, but it is an analytic function of $φ$ and $α$, which is convenient for derivatives and optimisation.  Wavelet theorists have studied the conditions on the set of amplitudes $A$ that allow any state to be expanded in this form, and the numerical stability of those expansions as the density of ampltiudes decreases towards the minimum for completeness.  This will be discussed later.

@@

The methods will consider to approximate $|ψ〉=|A〉d$ set $d$ to a linear function of $|ψ〉$.  We will sometimes denote this mapping $d=〈A⁺|ψ〉$, because we intend that $|ψ〉≈|A〉d=|A〉〈A⁺|ψ〉$, whence $〈A⁺|$ is a pseudoinverse of the expansion operator $|A〉$.  We can then define an error operator $E=\bigl(1-|A〉〈A⁺|\bigr)$, so that $|ψ〉-|A〉d=E|ψ〉$.  Because $〈A⁺|$ is linear, it is equivalent to a column vector of bras $〈A⁺|=\pmatrix{〈φ₁|&⋯&〈φ_R|}\T ,$
such that $d_r=〈φ_r|ψ〉$.  We will call $|φ_r〉$ the dual ket of the component $|α_r〉$.  Note that the dual kets are not normalised.

@@

There is another way to derive coefficients of the constant function, which might be more stable.  Suppose that $D$ is a subset of the real line, large enough that the tails of gaussians are neglibible outside it.  Let 
$$U=π^{-½}\pmatrix{e^{-(x-x₁)²}&⋯&e^{-(x-x_R)²}},$$
and $w$ be a column vector of $R$ real weights.  Then define a residual
$$r(w)=\int_D(Uw-1)².$$
The derivative of this with respect to the weights is
$$Dr(w)=\int_D 2(Uw-1)U=2\int_D U\T Uw-U.$$
(Note that, for row vectors $a$ and $c$, and $b$ a column vector, $abc=c\T ab$.)
Unfortunately,
$$\int (U\T U)_{ij}=π^{-1}\int e^{-(x-x_i)²-(x-x_j)²}\,dx
	={1\over\sqrt{2π}}e^{(x_i-x_j)²/2},$$
and $\int u_i=1$, so this is the identical problem again.

@@

To prevent the weights from oscillating in order to fit the edges of the samples, it is tempting to do a weighted least squares fit, which emphasises fitting the centre of the cluster over fitting the edges.  There are two ways to do this.  Firstly, instead of expanding the constant function $f(x)=1$, we can expand a peaked function such as $f(x)=e^{-x²}$, then divide the coefficient of $e^{-(x-x_r)²}$ by $e^{-(x_r)²}$ to get a weight.  This mimics the expansions we need to do in practice, where the sample points cover the region where the expanded function has large values.  The residual in this case is
$${\cal E} = \int\left(e^{-x²}-∑_rd_re^{(x-x_r)²}\right)²;$$
the only change to the normal equations is that the right hand side $\pmatrix{1&⋯&1}\T $ is replaced by a constant multiple of 
$$b=\pmatrix{e^{-x₁²/2}&⋯&e^{-x_R²/2}}.\tag\rhs$$
For now, I'm ignoring factors of 2, $\sqrt π $ and so on: how the weights oscillate is more interesting than their size.  This seems to work: with random sample points, the weights still oscillate, but remain within an order of magnitude or two, where the weights from expanding the constant function vary by a factor of a million.
